Learning how to act: making good decisions with machine learning

### 2017-08-27 thesis/thesis_map.tex
rough restructure complete

@@

\section*{Questions}
\begin{itemize}
\item Are there any problematic counterfactuals being used in defining bandit problems
\item in the Schölkopf causal fairness paper - is there an issue with the use of twin networks. Could it be fixed with SWIGs?
\item Contextual bandits are solving this argmax h problem. Should causal inference actually be solving the same thing?
\item what is the connection between which variables should be included when doing estimating causal effects from observational data and stratification in randomized experimental design?
\item I don't understand why covariate shift is not supposed to be an issue in Bayesian models. 
\item What does conditioning mean in the Bayesian sense? - can imposing a prior which is true with respect to that variable, increase bias due to confounding? (In the same way that in the pnemonia case, the coefficient for Asthma should be set to zero after the model has been fit, not before.

\item What are the nessessary and sufficient conditioins for decision making? How much about a the distribution of rewards for different arms must I know in order to select one (under different assumptions about the loss function). Ie when is the mean sufficient? What about the mean and the variance? - when can we take one set of results and compute something about a different loss function?

\item Does the way we define supervised learning and loss fuctions imply an additive noise model? Why would we expect this to hold - why don't we run into more problems?

@@

There are two ways in which an association detected in a dataset may be non-causal. The first is that the variables concerned may not be related at all, and the association has arisen by chance in that data sample. Given a finite data on enough variables, there is a high probability of finding some that appear correlated even though they are completely unrelated. For example, based on data from the years 1999 to 2009, the age of miss America is strongly correlated with the number of murders  (in the US) by steam, hot vapours and hot objects \citep{Vigen2015}. We would not expect this relationship to hold if we were to look at a new sample of data. This form of spurious correlation also has serious repercussions. It lies at the heart of major problems with the process of scientific research: researchers are incentivised to detect effects and thus to explore many possible paths in the process of analysing data and studies that fail to find an effect are less likely to be published. As a consequnce, the likelyhood that reported effects have arisen by chance is underestimated, leading to the conclusion that \quotes{most published scientific results are false} \citep{Ioannidis2005}. This issue is also highlighted by recent crises in replication \citep{OpenScienceCollaboration2015}. This issue can be ameliorated by getting more data and by seperating learning models from evaluating their performance, for example by evaluating models on a stict hold-out set or on the extent to which their results can be replicated.

However, a strong association, observed in multiple indpendent studies may still not be causal. The correlation can arise because both variables are consequences of some other, unmeasured factor. For example, the reading ability of children under twelve is strongy correlated with their height, because older children are taller and can read better. However height is not a cause of reading ability because if we were to intervene to increase a child's height, for example by giving them growth hormones, we would not expect dramatic improvements in their reading. Similarly, extra tutoring in reading will not make a child grow taller. This problem is fundamentally different to the issue of spurious correlations arising by chance in finite datasets. Obtaining more (even infinitely many more) samples without directly intervening in the system to manipulate the variables does not allow us to seperate causation from correlation.

The key distinction between a real, but non-causal, association and a causal relationship is in what happens if we intervene in the system and change one of the variables. In this thesis, I take an interventionalist viewpoint of causality: any model or approach designed to predict the outcome of intervening in a system is causal. This viewpoint captures the types of questions that motivate this thesis. How can we change the way we do things to obtain better outcomes.

Causality is often liked to explaination; understanding how and why things happen. I view explaination in terms of compression and generalisabilty: how much information about the world a model can capture. This creates a hierachy in the degree to which models are explainitory, rather than a simple binary distinction. A standard predictive model encodes all the information we need to predict some output given inputs provided the system generating the data does not change. A high-level causal model might be able to predict the outcome of a specific intervention holding all else fixed. More detailed causal models could predict the outcome for a wide range of combiations of interventions conditional on a range of contexts. By considering conditional interventions within our definition of causal questions we also capture mediation: the study of through which pathways one variable causes another \cite{vanderweele2015explanation}. Finally, a model that can distil how elements interact to mathematical equatons like Newton's laws can be used to predict what will happen in an astounding range of settings, including many we have never previously observed \footnote{Although Newton's laws are not fully general, as they break down for extremely massive objects}.

\citet{Gelman2010,gelman2013ask} make a distinction between forward causal inference, the types of \quotes{What if} questions I focus on in this thesis, and reverse causal questions, asking why something occurs. The former aims to identify the effect of a known cause. The latter can be viewed as identiying causes of an effect. They regard forward causal inference as well defined within the counterfactual and graphical model frameworks for causal inference, which we describe in chapter \ref{chap:causal_models}. However, state that \quotes{A reverse causal question does not in general have a well-defined answer, even in a setting where all possible data are made available.} I view this as overly pessimistic, depending on how \quotes{all possible data} is defined. The goal of identifying the causes of an effect can be formalised within the graphical causal model framework. Solving this problem is certainly much more challanging than identifying the effect of a specific intervention on given outcome, since it requires us to test or infer the effect of interventions on many different variables. These practical difficulties may well be overwealming, particularly in fields such as social science and economics where datasets are relatively small, systems are complex, the variables are difficult to directly manipulate and even relatively simple "What if" questions are hard to resolve conclusively. However, this does not mean that the problem of identifying causes of effects is ill-posed in principle. It can be viewed as a form of causal discovery: the attempt to learn the structure of the causal relationships between variables, on which there is a rich liturature which we review briefly in section \ref{sec:causal-discovery}.
\todo{reference a review paper on causal discovery here}

There has traditionally been a large gap between researches in machine learning who focus on prediction, use largly non-interpretable models and researches in statistics, social science and economics who (at least implicitly) aim to answer causal questions and tend to use heavily theory driven models. However, there is relatively little awareness, particularly within the machine learning and data science communities, of what constitutes a causal problem and the implications of for the training and evaluation models. In the next section we emphasise the subtley that can exist in determining if a problem is causal by examiming some typical examples.

@@

Machine learning is in the midst of a boom, driven by the availability of large datasets and the computation resources to process them. Machine learning techniques are being applied a huge range of problems, in both industry and achademia. The following examles are inteded to capture the bredth of problems that machine learning algorithms are actively beeing applied to. Which, if any, of these problems require causal inference?

@@

The question is disingenuous because I have not posed the problems in sufficient detail to determine if causality is an important considertion. In particular, I failed to specify how any model we might build would be used: what actions would be taken in response to its predictions.

@@

This example also highlights a connection between causal models and transparancy in machine learning. If we are using a non-causal model to make decision effecting people, there will be a trade-off between the performance and transparancy of the model; not because the requirement for transparancy restricts us to simple models but because revealing how the model works allows people to change their behaviour to game it.

What about predicting the risk of death in patients with pneumonia? Suppose the goal is to build a model to decide who should be treated in hospital and who can be sent home with antibiotics. If we assume that in hospital treatment is more effective for serious cases, this seems like a straightforward prediction problem. It is not. Depending on how the decision to admit was previously made and what features are included (or omitted) in the model, the relationship between those features and the outcome may change if the model is used to make admition decisons. \citet{Caruana2015} found exactly this effect in a real data set. The model learnt that people suffering asthma were \emph{less} likely to die from pneumonia. They realised this was because doctors were treating such patients very aggressively, thus actually lowering their risk. The issue is not with model, it performed very well at the task for which it was trained, which is to predict wo would be likely to die under the original admition and treatment protocols. However, using it to decide how to \emph{change} these protocols could kill. The actual question of interest in this case is what happens to patients with charachteristics $X$ if we assign them treatment according to decision rule $Z$.

@@

Playing Go (and other games) is a case with some subtleties. At every turn, the AI agent has a number of actions available. The board state following each action is deterministic and given by the rules of the game. The agent can apply supervised machine learning, based on millions of previous games, to estimate the probability that each of these reachable board states will lead to a win \footnote{This is a challenging pattern recognition problem. There are around $2 \times 10^{170}$ legal board positions in Go, (\citep{tromp2016number}), so the algorithm cannot simply memorise the proportion of times each state leads to a win, it must identify higher level features of the board state that are associated with winning.}. Supervised learning can also be applied to learn a policy $\P{a|s}$, the probability of selecting action $a$ given board state $s$. This allows the agent to estimate the likelyhood of winning from a given starting state by simulating (many times) the remainder of the game, drawing actions from $\P{a|s}$ for both players. Google's Alpha Go, which in May 2017 beat the then strongest human player \citep{Mozur2017}, incorporates a compbination of these approaches \citep{silver2016mastering}. The supervised learning was enhanced by having the agent play (varients) of itself many times, so that its estimates of value for each board state and of the likelyhood an opponent will play as given move are based on a combination of replicating the way humans play and on the moves that led to a win when playing itself.

The problem of playing go is causal from the interventionalist perspective. The agent wishes to learn the probability of a win given an action they take. However, there are some special characteristics to the go problem that make it amenable to a primarily supervised learning approach. The actions the agent has to explore are the sames ones as human players explored to generate the trainging data, and both have the same objective - to win the game. In addition, the state of the board encapsulates all the information relevent to selecting a move. These factors make it reasonable to conclude that selecting moves with an algorithm will not change the value of a board state or the probability of given move by the opponent by a sufficiently large margin to invalidate the training data.

Having considered these examples we can now identify some general aspects of problems that require causal (as opposed to purely predicive) inference. A predictive model may be sufficient if, given the variable(s) being predicted, it is clear which action is optimal and if selecting actions on the basis of the model does not change the mapping from features to outcomes. The second requirement is particularly difficult to satisfy when an algorithm is making important decisions effecting individual people. Think about problems like credit scoring and parole decisons. There are strong ethical grounds for demanding transparancy, but if the goals of society and the individuals are not perfectly aligned and there is any posibility that people can manipulate features independtly of the outcome, there will be a conflict between model accuracy and transparancy. It is rare to build a model without any intent to make some kind of decion on the basis of its results. Thus, I argue we should assume a causal model is required until we can justify otherwise.

@@

-------------------

In chapter \ref{chap:interventionalist_view} I describe the interventionalist viewpoint on learning the outcome of actions.

I review the key problems and results within the bandit liturature

highlight some interesting connections between existing strands of work on bandits and causal inference.

-------------------

Causal questions are everywhere. Statistics, economics, social science, artificial intelligence and machine learning. Techniques and for addressing such problems have developed in parralel within many diciplines. These techniques can usefully be categorised into two broad approaches, reinforcement learning and observational causal inference.

In reinforcement learning, under which we include traditional randomised experiments, we learn the outcome of actions by taking them. We take the role of an agent, capable of intervening in the system, and aim to develop algorithms that allow the agent to select actions optimally with respect to some goal. A particular strand of research within reinforcement learning are multi-armed bandit problems. They describe settings in which there are a set of available actions, the agent repeately decides which to select and then observes the outcome of the chosen action. They capture problems such as a doctor deciding which treatment to prescribe to a patient or a search engine selecting which advertisement to display to a user, where the agent faces the same set of choices repeatedly and is able to assess the value of outcome of the actions.

The approach of learning the outcome of an action by taking it has plays a key role in advancing our knoweledge of the world. However, we frequently have access to large bodies of data that have been collected from a system in which we did not have any control over what actions were taken, or perfect knowledge of the basis on which those actions were chosen. Estimating the affect of an action from such observational data sets is the problem addressed by observational causal inference.

Observational causal inference can be viewed as a form of transfer learning. The goal is to leverage data obtained from one system, the system we have observed, to estimate key characteristics of another, the system after we select an action via some policy that may differ from the process driving which actions occur in the orignial system. This is impossible without some assumptions about how the the orignal system and the system after intervention are related to one-another. The key to observational inference is a framework that models how actions change the state of the world and thus allows us to map information collected in one setting to another.

@@

The goals of this thesis are: to clarify the distinction between prediction and causal inference and highlight when and how we can leverage advances in prediction to improve causal inference and to draw together the key approaches to solving causal problems, from both the interventionalist and observational viewpoints. In chapter two, I describe the existing frameworks for formalising causality developed withing economics, statistics and machine learning and discuss how they relate to one another. Chapter three reviews the key approaches to causal inference from observational data. Chapter four highlights the interventionalist viewpoint, covering traditional randomised experiments and the closely related multi-armed bandit problems and highlights connections between the algorithms applied to bandit problems and those for estimating causal effects from observational data.

In chapter five, I develop a framework that unifies the causal graphical model approach for inference in observational settings with the sequential experimental approach encapsulated by multi-armed bandits. The causal models allow us to represent our knowledge of how variables are related to one-another in a very natural way and induces an interesting and novel form of structure between the different actions modelled in the bandit problem. I develop a new algorithm that can exploit this structure as a first step towards a unified approach to decision making under uncertainty.

Chapters two to four predominantly review the existing liturature with the goal of making connections between the different viewpoints and approaches. They do not contain novel technical results but present the material in a unified way. Chapter five is a novel contribution. It formally connect causal graphical models and multi-armed bandit problems, demonstrates this leads to a form of structure that had not previously been analysed within the bandit liturature and describes an algorithm that can leverage this structure to make better decisons.

@@

The goal of causal inference is to learn the effect of taking an action. We can do this directly via experimental approaches, however any given agent only has a limited capacity to manipulate the world. We are generating and storing data on almost every aspect of our lives at an unprecedented rate. As we incorporate sensors and robotics into our cities, homes, cars, everyday products and even our bodies, the breadth and scale of this data will only increase. However, only a tiny fraction of this data will be generated in a controlled way with the specific goal of answering a single question. An agent that can only learn from data when it had explicit control (or perfect knowledge of) the process by which that data was generated will be severly limted. This makes it critical that we develop effective methods that us to predict the outcome of an intervention in some system from observing it. This is the problem of observational causal inference.

@@

As previusly mentioned, observational causal inference is

Observational causal inference aims to infer the outcome of an intervention in some system from data obtained by observing (but not intervening in) it. As previuosly mentioned, this is a form of transfer learning; we need to infer properties of the system post-intervention from observations of the system pre-intervention. Doing this requires some assumptions about how these two systems are related, or other words, a way of describing actions and how we anticipate the system will respond to them. Three key approaches have emerged: counterfactuals, structural equation models and causal Bayesian networks.

counterfactuals

MOTIVATE WHY I AM DESCRIBING THEM ALL SUMMARISE WHAT EACH OFFERS.

In this chapter we will examine the problems these approaches allow us to solve, the assumptions they rely on and how they differ. We will also use them to describe the following simplified examples. The aim is to demonstrate the notations and formalisms needed to tackle more interesting problems later on.

@@

\begin{figure}
\centering
\begin{subfigure}[t]{0.15\textwidth}
\centering
\caption{}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm, thick,main node/.style={observed}]
 %nodes
\node[main node](1){$X_{1}$};
\node[main node, below=of 1](2){$X_{2}$};
\node[main node, below=of 2](3){$X_{3}$};
\path[every node/.style={font=\sffamily\small}]
    (1) edge (2)
    (2) edge (3);
\end{tikzpicture}
\end{subfigure}
\begin{subfigure}[t]{0.15\textwidth}
\centering
\caption{}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm, thick,main node/.style={observed}]
 %nodes
\node[main node](1){$X_{1}$};
\node[main node, below=of 1](2){$X_{2}$};
\node[main node, below=of 2](3){$X_{3}$};
\path[every node/.style={font=\sffamily\small}]
    (3) edge (2)
    (2) edge (1);
\end{tikzpicture}
\end{subfigure}
\begin{subfigure}[t]{0.3\textwidth}
\centering
\caption{}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm, thick,main node/.style={observed}]
 %nodes
\node[main node](1){$X_{2}$};
\node[main node, below left=of 1](2){$X_{1}$};
\node[main node, below right=of 1](3){$X_{3}$};
\path[every node/.style={font=\sffamily\small}]
    (1) edge (2) edge (3);
\end{tikzpicture}
\end{subfigure}
\begin{subfigure}[t]{0.3\textwidth}
\centering
\caption{}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm, thick,main node/.style={observed}]
 %nodes
\node[main node](1){$X_{2}$};
\node[main node, above left=of 1](2){$X_{1}$};
\node[main node, above right=of 1](3){$X_{3}$};
\path[every node/.style={font=\sffamily\small}]
    (2) edge (1)
    (3) edge (1)
    (2) edge (3);
\end{tikzpicture}
\end{subfigure}
\caption{Some valid Bayesian networks for a distribution $P$ over $(X_1,X_2,X_3)$ in which $X_3$ is conditionally independent of $X_1$ given $X_2$, denoted $X_3 \ci X_1 | X_2$. Graphs (a), (b) and (c) are all a \emph{perfect map} for $P$ as the graphical stucture implies exactly the same set of independencies exhibitied by the distribution. Graph (d), like figure \ref{fig:bayesnet} does not imply any conditional independencies, and is thus a valid (but not very useful) Bayesian network representation for any distribution over three variables.}
\label{fig:bayesnet2}
\end{figure}

@@

Causal Bayesian networks are Bayesian networks, so results that apply to Bayesian networks carry directly across: the local Markov property states that variables are independent of their non-effects given their direct causes. The global Markov property and d-separation also hold in causal networks. D-speration, which characterises which conditional independencies must hold in any distribution that can represented by a given Bayesian network $G$, is key to many important results and algorithms for causal inference. We include a brief introduction in section \ref{sec:dsep}.

>>>>

\setcounter{secnumdepth}{4}

@@

\section*{Questions}
\begin{itemize}
\item Are there any problematic counterfactuals being used in defining bandit problems
\item in the Schölkopf causal fairness paper - is there an issue with the use of twin networks. Could it be fixed with SWIGs?
\item Contextual bandits are solving this argmax h problem. Should causal inference actually be solving the same thing?
\item what is the connection between which variables should be included when doing estimating causal effects from observational data and stratification in randomised experimental design?
\item I don't understand why covariate shift is not supposed to be an issue in Bayesian models. 
\item What does conditioning mean in the Bayesian sense? - can imposing a prior which is true with respect to that variable, increase bias due to confounding? (In the same way that in the pneumonia case, the coefficient for Asthma should be set to zero after the model has been fit, not before.

\item What are the necessary and sufficient conditions for decision making? How much about a the distribution of rewards for different arms must I know in order to select one (under different assumptions about the loss function). Ie when is the mean sufficient? What about the mean and the variance? - when can we take one set of results and compute something about a different loss function?

\item Does the way we define supervised learning and loss functions imply an additive noise model? Why would we expect this to hold - why don't we run into more problems?

@@

There are two ways in which an association detected in a data set may be non-causal. The first is that the variables concerned may not be related at all, and the association has arisen by chance in that data sample. Given a finite data on enough variables, there is a high probability of finding some that appear correlated even though they are completely unrelated. For example, based on data from the years 1999 to 2009, the age of miss America is strongly correlated with the number of murders  (in the US) by steam, hot vapours and hot objects \citep{Vigen2015}. We would not expect this relationship to hold if we were to look at a new sample of data. This form of spurious correlation also has serious repercussions. It lies at the heart of major problems with the process of scientific research: researchers are incentivised to detect effects and thus to explore many possible paths in the process of analysing data and studies that fail to find an effect are less likely to be published. As a consequence, the likelyhood that reported effects have arisen by chance is underestimated, leading to the conclusion that \quotes{most published scientific results are false} \citep{Ioannidis2005}. This issue is also highlighted by recent crises in replication \citep{OpenScienceCollaboration2015}. This issue can be ameliorated by getting more data and by separating learning models from evaluating their performance, for example by evaluating models on a strict hold-out set or on the extent to which their results can be replicated.

However, a strong association, observed in multiple independent studies may still not be causal. The correlation can arise because both variables are consequences of some other, unmeasured factor. For example, the reading ability of children under twelve is strongly correlated with their height, because older children are taller and can read better. However height is not a cause of reading ability because if we were to intervene to increase a child's height, for example by giving them growth hormones, we would not expect dramatic improvements in their reading. Similarly, extra tutoring in reading will not make a child grow taller. This problem is fundamentally different to the issue of spurious correlations arising by chance in finite data sets. Obtaining more (even infinitely many more) samples without directly intervening in the system to manipulate the variables does not allow us to separate causation from correlation.

The key distinction between a real, but non-causal, association and a causal relationship is in what happens if we intervene in the system and change one of the variables. In this thesis, I take an interventionist viewpoint of causality: any model or approach designed to predict the outcome of intervening in a system is causal. This viewpoint captures the types of questions that motivate this thesis. How can we change the way we do things to obtain better outcomes.

Causality is often liked to explanation; understanding how and why things happen. I view explanation in terms of compression and generalisabilty: how much information about the world a model can capture. This creates a hierarchy in the degree to which models are explanatory, rather than a simple binary distinction. A standard predictive model encodes all the information we need to predict some output given inputs provided the system generating the data does not change. A high-level causal model might be able to predict the outcome of a specific intervention holding all else fixed. More detailed causal models could predict the outcome for a wide range of combinations of interventions conditional on a range of contexts. By considering conditional interventions within our definition of causal questions we also capture mediation: the study of through which pathways one variable causes another \cite{vanderweele2015explanation}. Finally, a model that can distil how elements interact to mathematical equations like Newton's laws can be used to predict what will happen in an astounding range of settings, including many we have never previously observed \footnote{Although Newton's laws are not fully general, as they break down for extremely massive objects}.

\citet{Gelman2010,gelman2013ask} make a distinction between forward causal inference, the types of \quotes{What if} questions I focus on in this thesis, and reverse causal questions, asking why something occurs. The former aims to identify the effect of a known cause. The latter can be viewed as identifying causes of an effect. They regard forward causal inference as well defined within the counterfactual and graphical model frameworks for causal inference, which we describe in chapter \ref{chap:causal_models}. However, state that \quotes{A reverse causal question does not in general have a well-defined answer, even in a setting where all possible data are made available.} I view this as overly pessimistic, depending on how \quotes{all possible data} is defined. The goal of identifying the causes of an effect can be formalised within the graphical causal model framework. Solving this problem is certainly much more challenging than identifying the effect of a specific intervention on given outcome, since it requires us to test or infer the effect of interventions on many different variables. These practical difficulties may well be overwhelming, particularly in fields such as social science and economics where data sets are relatively small, systems are complex, the variables are difficult to directly manipulate and even relatively simple "What if" questions are hard to resolve conclusively. However, this does not mean that the problem of identifying causes of effects is ill-posed in principle. It can be viewed as a form of causal discovery: the attempt to learn the structure of the causal relationships between variables, on which there is a rich literature which we review briefly in section \ref{sec:causal-discovery}.
\todo{reference a review paper on causal discovery here}

There has traditionally been a large gap between researches in machine learning who focus on prediction, use largely non-interpretable models and researches in statistics, social science and economics who (at least implicitly) aim to answer causal questions and tend to use heavily theory driven models. However, there is relatively little awareness, particularly within the machine learning and data science communities, of what constitutes a causal problem and the implications of for the training and evaluation models. In the next section we emphasise the subtly that can exist in determining if a problem is causal by examining some typical examples.

@@

Machine learning is in the midst of a boom, driven by the availability of large data sets and the computation resources to process them. Machine learning techniques are being applied a huge range of problems, in both industry and academia. The following examples are intended to capture the breadth of problems that machine learning algorithms are actively being applied to. Which, if any, of these problems require causal inference?

@@

The question is disingenuous because I have not posed the problems in sufficient detail to determine if causality is an important consideration. In particular, I failed to specify how any model we might build would be used: what actions would be taken in response to its predictions.

@@

This example also highlights a connection between causal models and transparency in machine learning. If we are using a non-causal model to make decision effecting people, there will be a trade-off between the performance and transparency of the model; not because the requirement for transparency restricts us to simple models but because revealing how the model works allows people to change their behaviour to game it.

What about predicting the risk of death in patients with pneumonia? Suppose the goal is to build a model to decide who should be treated in hospital and who can be sent home with antibiotics. If we assume that in hospital treatment is more effective for serious cases, this seems like a straightforward prediction problem. It is not. Depending on how the decision to admit was previously made and what features are included (or omitted) in the model, the relationship between those features and the outcome may change if the model is used to make admission decisions. \citet{Caruana2015} found exactly this effect in a real data set. The model learnt that people suffering asthma were \emph{less} likely to die from pneumonia. They realised this was because doctors were treating such patients very aggressively, thus actually lowering their risk. The issue is not with model, it performed very well at the task for which it was trained, which is to predict who would be likely to die under the original admission and treatment protocols. However, using it to decide how to \emph{change} these protocols could kill. The actual question of interest in this case is what happens to patients with characteristics $X$ if we assign them treatment according to decision rule $Z$.

@@

Playing Go (and other games) is a case with some subtleties. At every turn, the AI agent has a number of actions available. The board state following each action is deterministic and given by the rules of the game. The agent can apply supervised machine learning, based on millions of previous games, to estimate the probability that each of these reachable board states will lead to a win \footnote{This is a challenging pattern recognition problem. There are around $2 \times 10^{170}$ legal board positions in Go, (\citep{tromp2016number}), so the algorithm cannot simply memorise the proportion of times each state leads to a win, it must identify higher level features of the board state that are associated with winning.}. Supervised learning can also be applied to learn a policy $\P{a|s}$, the probability of selecting action $a$ given board state $s$. This allows the agent to estimate the likelyhood of winning from a given starting state by simulating (many times) the remainder of the game, drawing actions from $\P{a|s}$ for both players. Google's Alpha Go, which in May 2017 beat the then strongest human player \citep{Mozur2017}, incorporates a combination of these approaches \citep{silver2016mastering}. The supervised learning was enhanced by having the agent play (variants) of itself many times, so that its estimates of value for each board state and of the likelyhood an opponent will play as given move are based on a combination of replicating the way humans play and on the moves that led to a win when playing itself.

The problem of playing go is causal from the interventionalist perspective. The agent wishes to learn the probability of a win given an action they take. However, there are some special characteristics to the go problem that make it amenable to a primarily supervised learning approach. The actions the agent has to explore are the sames ones as human players explored to generate the training data, and both have the same objective - to win the game. In addition, the state of the board encapsulates all the information relevant to selecting a move. These factors make it reasonable to conclude that selecting moves with an algorithm will not change the value of a board state or the probability of given move by the opponent by a sufficiently large margin to invalidate the training data.

Having considered these examples we can now identify some general aspects of problems that require causal (as opposed to purely predictive) inference. A predictive model may be sufficient if, given the variable(s) being predicted, it is clear which action is optimal and if selecting actions on the basis of the model does not change the mapping from features to outcomes. The second requirement is particularly difficult to satisfy when an algorithm is making important decisions effecting individual people. Think about problems like credit scoring and parole decisions. There are strong ethical grounds for demanding transparency, but if the goals of society and the individuals are not perfectly aligned and there is any possibility that people can manipulate features independently of the outcome, there will be a conflict between model accuracy and transparency. It is rare to build a model without any intent to make some kind of decision on the basis of its results. Thus, I argue we should assume a causal model is required until we can justify otherwise.

@@

As we have shown, problems involving causality are ubiquitous in many fields. As a result, techniques for addressing them have developed in parallel within many disciplines, including statistics, economics, social science, epidemiology and machine learning. Although the focus and terminology can differ substantially between fields these techniques all address the underlying goal of estimating the effect of or optimally selecting interventions in some system. Methods for learning about interventions can be usefully categorised into two broad approaches, reinforcement learning and observational causal inference.

In reinforcement learning, under which we include traditional randomised experiments, we learn the outcome of actions by taking them. We take the role of an agent, capable of intervening in the system, and aim to develop algorithms that allow the agent to select actions optimally with respect to some goal. A particular strand of research within reinforcement learning are multi-armed bandit problems. They describe settings in which there are a set of available actions, the agent repeatedly decides which to select and then observes the outcome of the chosen action. They capture problems such as a doctor deciding which treatment to prescribe to a patient or a search engine selecting which advertisement to display to a user, where the agent faces the same set of choices repeatedly and is able to assess the value of outcome of the actions.

The approach of learning the outcome of an action by taking it has plays a key role in advancing our knowledge of the world. However, we frequently have access to large bodies of data that have been collected from a system in which we did not have any control over what actions were taken, or perfect knowledge of the basis on which those actions were chosen. Estimating the affect of an action from such observational data sets is the problem addressed by observational causal inference. Observational causal inference can be viewed as a form of transfer learning. The goal is to leverage data obtained from one system, the system we have observed, to estimate key characteristics of another, the system after we select an action via some policy that may differ from the process driving which actions occur in the original system. This is impossible without some assumptions about how the the original system and the system after intervention are related to one-another. The key to observational inference is to model how actions change the state of the world in such a way that we can map information collected in one setting to another.

@@

The primary goal of this thesis is to draw together the key approaches to solving causal problems from both the interventionalist and observational viewpoints.

In chapter \ref{chap:causality} I describe key frameworks for causal inference from observational data and discuss how the relate to one-anther and review how they can be leveraged to answer causal queries. Chapter \ref{chap:interventionalist_view} covers the interventionalist viewpoint, including traditional randomised experiments and the closely related multi-armed bandit problems. It describes the key problems and results within the bandit literature and highlights some interesting connections between existing strands of work on bandits and causal inference. In chapter \ref{chap:causal-bandit}, I develop a framework that unifies the causal graphical model approach for inference in observational settings with the sequential experimental approach encapsulated by multi-armed bandits. The causal models allow us to represent our knowledge of how variables are related to one-another in a very natural way and induces an interesting and novel form of structure between the different actions modelled in the bandit problem. I develop a new algorithm that can exploit this structure as a first step towards a unified approach to decision making under uncertainty.

Chapters \ref{chap:causality} and \ref{chap:interventionalist_view} predominantly review the existing literature with the goal of highlighting connections between the different viewpoints and approaches. They do not contain novel technical results but present the material in a unified way. Chapter \ref{chap:causal-bandit} is a novel contribution. It formally connect causal graphical models and multi-armed bandit problems, demonstrates this leads to a form of structure that had not previously been analysed within the bandit literature and describes an algorithm that can leverage this structure to make better decisions.

@@

The goal of causal inference is to learn the effect of taking an action. We can do this directly via experimental approaches, however any given agent only has a limited capacity to manipulate the world. We are generating and storing data on almost every aspect of our lives at an unprecedented rate. As we incorporate sensors and robotics into our cities, homes, cars, everyday products and even our bodies, the breadth and scale of this data will only increase. However, only a tiny fraction of this data will be generated in a controlled way with the specific goal of answering a single question. An agent that can only learn from data when it had explicit control (or perfect knowledge of) the process by which that data was generated will be severely limited. This makes it critical that we develop effective methods that us to predict the outcome of an intervention in some system from observing it. This is the problem of observational causal inference.

@@

Observational causal inference aims to infer the outcome of an intervention in some system from data obtained by observing (but not intervening in) it. As previously mentioned, this is a form of transfer learning; we need to infer properties of the system post-intervention from observations of the system pre-intervention. Mapping properties from one system to another requires some assumptions about how these two systems are related, or other words, a way of describing actions and how we anticipate a system will respond to them. Three key approaches have emerged: counterfactuals, structural equation models and causal Bayesian networks.

Counterfactuals \citep{Rubin1974} were developed from the starting point of generalising from randomised trials to less controlled settings. They describe causal effects in terms of differences between counterfactual variables, what would happen if we took one action verses what would happen if we took another. Counterfactual assertions can be expressed very naturally in human languages and are quite prevalent in  every day conversations;\quotes{if I had worked harder I would have got better grades}, \quotes{she would have been much sicker if she hadn't taken antibiotics}. Structural equation models have been developed and applied primarily within economics and related disciplines and can be seen as an attempt to capture key aspects of the behaviour of people with mathematics. Questions around designing policies or interventions play a central role in economics. Thus they have transformed simultaneous equations into a powerful (though somewhat poorly formalised) framework and associated set of methods for estimating causal effects. Causal Bayesian networks \citep{Pearl2000} are a more recent development and arise from the addition of a fundamental assumption about the meaning of a link to Bayesian networks. They inherit and leverage the way Bayesian networks encode conditional independencies between variables to localise the impact of an intervention in a system in a way that allows formalisation of conditions under causal effects can be inferred from observational data.

An understanding of causal Bayesian networks and their properties (in particular the do calculus, see section \ref{sec:do_calculus}) is sufficient to appreciate my main technical contributions in chapter \ref{chap:causal-bandit}, as well as the importance of formal causal reasoning in Bayesian inference that I highlight in section  \ref{sec:bayesian_causality}. However, the literature on causal inference techniques remains spit between the different frameworks. Much of the recent work on estimating causal effects within machine learning, as well as widely used methodologies such as propensity scores, are described using the counterfactual framework. Methods developed within economics, in particular instrumental variable based approaches, or those requiring parametric or functional assumptions are often based around structural equation models. This makes it worthwhile for researchers interested in causality to develop an understanding off all these viewpoints.

In the next sections we describe causal Bayesian networks, counterfactuals and structural equation models: the problems they he problems they allow us to solve, the assumptions they rely on and how they differ. We hope that by describing all three frameworks, how they relate to one-another, and when they can be viewed as equivalent, we will make it easier for researchers familiar with one framework to understand the others and to transfer ideas and techniques between them. However, sections \ref{sec:SEM} (Structural Equation models), and \ref{sec:unifying_causal_models} (Unifying the models) are not crucial to understanding my technical contributions and may be safely skipped by uninterested readers. In order to demonstrate the notation and formalisms each framework provides we will use them to describe the following simple examples.

@@

\begin{figure}
\centering
\begin{subfigure}[t]{0.15\textwidth}
\centering
\caption{}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm, thick,main node/.style={observed}]
 %nodes
\node[main node](1){$X_{1}$};
\node[main node, below=of 1](2){$X_{2}$};
\node[main node, below=of 2](3){$X_{3}$};
\path[every node/.style={font=\sffamily\small}]
    (1) edge (2)
    (2) edge (3);
\end{tikzpicture}
\end{subfigure}
\begin{subfigure}[t]{0.15\textwidth}
\centering
\caption{}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm, thick,main node/.style={observed}]
 %nodes
\node[main node](1){$X_{1}$};
\node[main node, below=of 1](2){$X_{2}$};
\node[main node, below=of 2](3){$X_{3}$};
\path[every node/.style={font=\sffamily\small}]
    (3) edge (2)
    (2) edge (1);
\end{tikzpicture}
\end{subfigure}
\begin{subfigure}[t]{0.3\textwidth}
\centering
\caption{}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm, thick,main node/.style={observed}]
 %nodes
\node[main node](1){$X_{2}$};
\node[main node, below left=of 1](2){$X_{1}$};
\node[main node, below right=of 1](3){$X_{3}$};
\path[every node/.style={font=\sffamily\small}]
    (1) edge (2) edge (3);
\end{tikzpicture}
\end{subfigure}
\begin{subfigure}[t]{0.3\textwidth}
\centering
\caption{}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm, thick,main node/.style={observed}]
 %nodes
\node[main node](1){$X_{2}$};
\node[main node, above left=of 1](2){$X_{1}$};
\node[main node, above right=of 1](3){$X_{3}$};
\path[every node/.style={font=\sffamily\small}]
    (2) edge (1)
    (3) edge (1)
    (2) edge (3);
\end{tikzpicture}
\end{subfigure}
\caption{Some valid Bayesian networks for a distribution $P$ over $(X_1,X_2,X_3)$ in which $X_3$ is conditionally independent of $X_1$ given $X_2$, denoted $X_3 \ci X_1 | X_2$. Graphs (a), (b) and (c) are all a \emph{perfect map} for $P$ as the graphical structure implies exactly the same set of independencies exhibited by the distribution. Graph (d), like figure \ref{fig:bayesnet} does not imply any conditional independencies, and is thus a valid (but not very useful) Bayesian network representation for any distribution over three variables.}
\label{fig:bayesnet2}
\end{figure}

@@
